---
title: "Week 03 Assignment"
output: html_notebook
---
For this week’s application assignment, please complete the following tasks using the data provided in clinic.xls (see Week04 folder)

1. Tabulate the execution times of each of the individual approaches for computing distance in Python (i.e., run the shared code on your computer, note the times, and tabulate them).

2. Next, replicate the for-loop based approach (the first one) and two different ways to make that version more efficient, in R. Profile these three approaches, and tabulate the results.

3. Based on the computational efficiency of implementations in Python and R, which one would you prefer? Based on a consideration of implementation (i.e., designing and implementing the code), which approach would you prefer? Taking both of these (run time and coding time), which approach would you prefer?

4. Identify and describe one or two other considerations, in addition to these two, in determining which of the two environments – Python or R – is preferable to you.

```{r}
install.packages("readxl")
```

```{r}
library(readxl)
```

```{r}
df <- read.csv("/Users/jacksparrow/Desktop/Assignment/clinics (1).csv", sep="|", header=TRUE)
head(df)
```
```{r}
library(readr)
library(geosphere)
```

```{r}
cat("Loading dataset...\n")
data <- read_delim("clinics (1).csv", delim = "|")
```
```{r}
# Validate and clean dataset
data <- data[!is.na(data$locLat) & !is.na(data$locLong), ]  
data$locLat <- as.numeric(data$locLat)
data$locLong <- as.numeric(data$locLong)
data <- data[data$locLat >= -90 & data$locLat <= 90 & data$locLong >= -180 & data$locLong <= 180, ]  # Keep valid coordinates
```

```{r}
# Approach 1: For-loop based distance calculation
compute_distances_for_loop <- function(data) {
  cat("Computing distances using for-loop...\n")
  distances <- c()
  for (i in 1:nrow(data)) {
    for (j in (i + 1):nrow(data)) {
      coord1 <- c(data$locLong[i], data$locLat[i])
      coord2 <- c(data$locLong[j], data$locLat[j])
      distance <- distHaversine(coord1, coord2) / 1000  # Convert to kilometers
      distances <- c(distances, distance)
    }
  }
  return(distances)
}
```

```{r}
# Measure execution time for for-loop approach
start_time <- Sys.time()
distances_for_loop <- compute_distances_for_loop(data)
end_time <- Sys.time()
loop_time <- end_time - start_time
```
```{r}
# Approach 2: Apply-based distance calculation
compute_distances_apply <- function(data) {
  cat("Computing distances using apply function...\n")
  coords <- data[, c("locLong", "locLat")]
  distances <- outer(1:nrow(coords), 1:nrow(coords), Vectorize(function(i, j) {
    if (i < j) {
      distHaversine(coords[i, ], coords[j, ]) / 1000  
    } else {
      NA
    }
  }))
  return(na.omit(as.vector(distances)))
}
```

```{r}
# Measure execution time for apply approach
start_time <- Sys.time()
distances_apply <- compute_distances_apply(data)
end_time <- Sys.time()
apply_time <- end_time - start_time
```

```{r}
#tabulate and print results
cat("tabulating results...\n")
results <- data.frame(
  Approach = c("For-loop", "Apply"),
  Time = c(loop_time, apply_time)
)
print(results)
```